---
title: "Supplementary Appendix: Complete Lists of Selected Markers"
subtitle: "QTL Mapping of Cotton Fibre Quality Traits"
author: "Shiyan Miao"
date: today
format:
  pdf:
    documentclass: article
    toc: true
    lot: true
    colorlinks: true
    number-sections: true
    number-depth: 2
    keep-tex: false
    geometry:
      - margin=1in
    fontsize: 10pt
execute:
  echo: false
  warning: false
  message: false
editor: visual
---

```{r setup}
library(tidyverse)
library(data.table)
library(glmnet)
library(knitr)
library(kableExtra)
library(scales)
library(irlba)

set.seed(2026)

knitr::opts_chunk$set(fig.align = "center")

geno_file  <- "Genotype_data.txt"
map_file   <- "Genetic_map.txt"
pheno_file <- "Phenotype_data.txt"
```

```{r load-and-process}
# ── Import ────────────────────────────────────────────────────────────────────
geno_raw  <- fread(geno_file, header = FALSE)
map_df    <- fread(map_file)
# Standardise all three column names positionally.
# map_df always has exactly 3 columns: marker ID, chromosome, position.
# This avoids any conflict with ggplot2::Position in tidyselect contexts.
names(map_df) <- c("SNP_ID", "Chromosome", "Pos_cM")
pheno_raw <- fread(pheno_file)

geno_df           <- as.data.frame(geno_raw)
colnames(geno_df) <- map_df$SNP_ID
geno_df$Line      <- seq_len(nrow(geno_df))

n_lines <- nrow(geno_df)
n_pheno <- nrow(pheno_raw)

pheno_df <- pheno_raw %>%
  mutate(
    RowID  = row_number(),
    Season = case_when(
      RowID <= n_lines       ~ "Season_1",
      RowID <= 2 * n_lines   ~ "Season_2",
      TRUE                   ~ "Season_3"
    ),
    Line = rep(seq_len(n_lines), times = n_pheno / n_lines)
  ) %>%
  relocate(Line, Season, RowID)

analysis_df <- pheno_df %>% left_join(geno_df, by = "Line")

trait_names <- c("Lint_percentage", "Seed_index",
                 "Lint_index",      "Seed_oil_content")

trait_labels <- c(
  Lint_percentage  = "Lint Percentage",
  Seed_index       = "Seed Index",
  Lint_index       = "Lint Index",
  Seed_oil_content = "Seed Oil Content"
)
```

```{r prepare-genotype}
# ── Genotype matrix: impute, filter, PCA ──────────────────────────────────────
X_geno       <- geno_df %>% select(-Line) %>% as.matrix()
mode(X_geno)  <- "numeric"

for (j in seq_len(ncol(X_geno))) {
  miss <- is.na(X_geno[, j])
  if (any(miss)) X_geno[miss, j] <- mean(X_geno[, j], na.rm = TRUE)
}

marker_keep  <- apply(X_geno, 2, function(x) var(x, na.rm = TRUE) > 0)
X_geno       <- X_geno[, marker_keep, drop = FALSE]
marker_names <- colnames(X_geno)

set.seed(123)
pca_fit <- prcomp_irlba(
  x = scale(X_geno, center = TRUE, scale = TRUE),
  n = 20, center = FALSE, scale. = FALSE
)

pc_line_df  <- as_tibble(pca_fit$x[, 1:5]) %>%
  setNames(paste0("PC", 1:5)) %>%
  mutate(Line = seq_len(n()))

analysis_df <- analysis_df %>%
  select(-any_of(paste0("PC", 1:5))) %>%
  left_join(pc_line_df, by = "Line") %>%
  mutate(Season = factor(Season))
```

```{r helper-functions}
prep_marker_matrix <- function(df, marker_names) {
  df <- as.data.frame(df)
  X  <- as.matrix(df[, marker_names, drop = FALSE])
  mode(X) <- "numeric"
  if (anyNA(X)) {
    col_means <- colMeans(X, na.rm = TRUE)
    idx       <- which(is.na(X), arr.ind = TRUE)
    X[idx]    <- col_means[idx[, 2]]
  }
  X
}

fast_single_marker_scan_cov <- function(df, trait, marker_names,
                                        covariates = "Season") {
  df          <- as.data.frame(df)
  needed_cols <- unique(c(trait, covariates, marker_names))
  df          <- df[, needed_cols, drop = FALSE]
  keep        <- complete.cases(df[, c(trait, covariates), drop = FALSE])
  df          <- df[keep, , drop = FALSE]

  y   <- df[[trait]]
  X   <- prep_marker_matrix(df, marker_names)
  C   <- model.matrix(
    as.formula(paste("~", paste(covariates, collapse = " + "))), data = df)
  qrC <- qr(C)

  y_res <- qr.resid(qrC, y)
  X_res <- qr.resid(qrC, X)
  Sxx   <- colSums(X_res^2)
  Sxy   <- as.vector(crossprod(X_res, y_res))
  Syy   <- sum(y_res^2)
  df_r  <- nrow(df) - qrC$rank - 1L
  valid <- Sxx > 1e-12

  beta <- rss <- se <- tval <- pval <- rep(NA_real_, length(marker_names))
  beta[valid]  <- Sxy[valid] / Sxx[valid]
  rss[valid]   <- pmax(Syy - Sxy[valid]^2 / Sxx[valid], 0)
  se[valid]    <- sqrt((rss[valid] / df_r) / Sxx[valid])
  tval[valid]  <- beta[valid] / se[valid]
  pval[valid]  <- 2 * pt(abs(tval[valid]), df = df_r, lower.tail = FALSE)

  tibble(trait = trait, SNP_ID = marker_names,
         estimate = beta, std.error = se,
         statistic = tval, p.value = pval)
}

add_multiple_testing <- function(scan_df, alpha = 0.05) {
  scan_df %>%
    group_by(trait) %>%
    arrange(p.value, .by_group = TRUE) %>%
    mutate(
      p_adj_bh   = p.adjust(p.value, method = "BH"),
      p_adj_bonf = p.adjust(p.value, method = "bonferroni"),
      sig_bh     = p_adj_bh   < alpha,
      sig_bonf   = p_adj_bonf < alpha
    ) %>%
    ungroup()
}

run_lasso_trait <- function(df, trait, marker_names,
                            covariates = c("Season","PC1","PC2","PC3","PC4","PC5"),
                            alpha = 1, nfolds = 10, seed = 123) {
  df   <- as.data.frame(df)
  keep <- complete.cases(df[, c(trait, covariates), drop = FALSE])
  df   <- df[keep, , drop = FALSE]

  y        <- df[[trait]]
  X_marker <- prep_marker_matrix(df, marker_names)
  X_cov    <- model.matrix(reformulate(covariates), data = df)[, -1, drop = FALSE]
  x_all    <- cbind(X_cov, X_marker)
  cov_names      <- colnames(X_cov)
  penalty_factor <- c(rep(0, ncol(X_cov)), rep(1, ncol(X_marker)))

  set.seed(seed)
  cvfit <- cv.glmnet(x_all, y, family = "gaussian", alpha = alpha,
                     nfolds = nfolds, standardize = TRUE,
                     intercept = TRUE, penalty.factor = penalty_factor)

  extract_markers <- function(s) {
    co  <- coef(cvfit, s = s)
    nz  <- setdiff(rownames(co)[as.vector(co != 0)], c("(Intercept)", cov_names))
    ef  <- as.vector(co[nz, ])
    tibble(SNP_ID = nz, coefficient = ef)
  }

  list(
    min = extract_markers("lambda.min"),
    `1se` = extract_markers("lambda.1se"),
    lambda_min = cvfit$lambda.min,
    lambda_1se = cvfit$lambda.1se
  )
}
```

```{r run-all-models}
# ── Single-marker scans ───────────────────────────────────────────────────────
scan_season <- lapply(trait_names, function(tr)
  fast_single_marker_scan_cov(analysis_df, tr, marker_names,
                              covariates = "Season")) %>%
  bind_rows() %>%
  left_join(map_df, by = "SNP_ID") %>%
  mutate(log10p = -log10(p.value)) %>%
  add_multiple_testing()

scan_pc <- lapply(trait_names, function(tr)
  fast_single_marker_scan_cov(analysis_df, tr, marker_names,
                              covariates = c("Season","PC1","PC2","PC3","PC4","PC5"))) %>%
  bind_rows() %>%
  left_join(map_df, by = "SNP_ID") %>%
  mutate(log10p = -log10(p.value)) %>%
  add_multiple_testing()

# ── Lasso: season + PCs ───────────────────────────────────────────────────────
lasso_pc <- lapply(trait_names, function(tr)
  run_lasso_trait(analysis_df, tr, marker_names,
                  covariates = c("Season","PC1","PC2","PC3","PC4","PC5"),
                  seed = 123))
names(lasso_pc) <- trait_names

# ── Lasso: season only ────────────────────────────────────────────────────────
lasso_season <- lapply(trait_names, function(tr)
  run_lasso_trait(analysis_df, tr, marker_names,
                  covariates = "Season", seed = 123))
names(lasso_season) <- trait_names
```

This appendix provides the complete lists of markers selected by all four model
specifications for each of the four cotton fibre quality traits. Tables are
organised by trait and then by model. Within each scan table, markers are ordered
by $p$-value. Within each lasso table, markers are ordered by the absolute value
of the estimated coefficient.

The four model specifications are:

1. **Season scan (BH)** — single-marker regression with season adjustment;
   Benjamini–Hochberg FDR threshold at 5 %.
2. **PC scan (BH)** — single-marker regression with season and PC1–5 adjustment;
   BH threshold at 5 %.
3. **Lasso + PC** — lasso with season and PC1–5 as unpenalised covariates;
   results reported for both $\lambda_{\min}$ and $\lambda_{1\text{se}}$.
4. **Lasso season-only** — lasso with season as the sole unpenalised covariate;
   results reported for both $\lambda_{\min}$ and $\lambda_{1\text{se}}$.

Map columns (chromosome and position) are included where available from the
genetic map file.

{{< pagebreak >}}

# Lint Percentage

## Season scan — BH-significant markers

```{r tbl-scan-season-lint}
#| label: tbl-scan-season-lint
#| tbl-cap: "BH-significant markers for Lint Percentage from the season-only single-marker scan, ordered by $p$-value."

scan_season %>%
  filter(trait == "Lint_percentage", sig_bh) %>%
  arrange(p.value) %>%
  select(SNP_ID, Chromosome, Pos_cM,
         Estimate = estimate, SE = std.error,
         `$t$` = statistic,
         `$p$-value` = p.value,
         `BH adj. $p$` = p_adj_bh) %>%
  knitr::kable(digits = c(0, 0, 2, 4, 4, 3, 2e-300, 3),
               format.args = list(scientific = TRUE)) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"),
                font_size = 8)
```

## PC-adjusted scan — BH-significant markers

```{r tbl-scan-pc-lint}
#| label: tbl-scan-pc-lint
#| tbl-cap: "BH-significant markers for Lint Percentage from the PC-adjusted single-marker scan, ordered by $p$-value."

scan_pc %>%
  filter(trait == "Lint_percentage", sig_bh) %>%
  arrange(p.value) %>%
  select(SNP_ID, Chromosome, Pos_cM,
         Estimate = estimate, SE = std.error,
         `$t$` = statistic,
         `$p$-value` = p.value,
         `BH adj. $p$` = p_adj_bh) %>%
  knitr::kable(digits = c(0, 0, 2, 4, 4, 3, 2e-300, 3),
               format.args = list(scientific = TRUE)) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"),
                font_size = 8)
```

## Lasso with PC adjustment

```{r tbl-lasso-pc-lint-min}
#| label: tbl-lasso-pc-lint-min
#| tbl-cap: "Markers selected by Lasso + PC for Lint Percentage under $\\lambda_{\\min}$, ordered by $|\\hat{\\beta}|$."

lasso_pc[["Lint_percentage"]]$min %>%
  left_join(map_df, by = "SNP_ID") %>%
  arrange(desc(abs(coefficient))) %>%
  select(SNP_ID, Chromosome, Pos_cM,
         `$\\hat{\\beta}$` = coefficient) %>%
  knitr::kable(digits = 4) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"),
                font_size = 8)
```

```{r tbl-lasso-pc-lint-1se}
#| label: tbl-lasso-pc-lint-1se
#| tbl-cap: "Markers selected by Lasso + PC for Lint Percentage under $\\lambda_{1\\text{se}}$, ordered by $|\\hat{\\beta}|$."

lasso_pc[["Lint_percentage"]]$`1se` %>%
  left_join(map_df, by = "SNP_ID") %>%
  arrange(desc(abs(coefficient))) %>%
  select(SNP_ID, Chromosome, Pos_cM,
         `$\\hat{\\beta}$` = coefficient) %>%
  knitr::kable(digits = 4) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"),
                font_size = 8)
```

## Lasso with season adjustment only

```{r tbl-lasso-season-lint-min}
#| label: tbl-lasso-season-lint-min
#| tbl-cap: "Markers selected by Lasso season-only for Lint Percentage under $\\lambda_{\\min}$."

lasso_season[["Lint_percentage"]]$min %>%
  left_join(map_df, by = "SNP_ID") %>%
  arrange(desc(abs(coefficient))) %>%
  select(SNP_ID, Chromosome, Pos_cM,
         `$\\hat{\\beta}$` = coefficient) %>%
  knitr::kable(digits = 4) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"),
                font_size = 8)
```

```{r tbl-lasso-season-lint-1se}
#| label: tbl-lasso-season-lint-1se
#| tbl-cap: "Markers selected by Lasso season-only for Lint Percentage under $\\lambda_{1\\text{se}}$."

lasso_season[["Lint_percentage"]]$`1se` %>%
  left_join(map_df, by = "SNP_ID") %>%
  arrange(desc(abs(coefficient))) %>%
  select(SNP_ID, Chromosome, Pos_cM,
         `$\\hat{\\beta}$` = coefficient) %>%
  knitr::kable(digits = 4) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"),
                font_size = 8)
```

{{< pagebreak >}}

# Seed Index

## Season scan — BH-significant markers

```{r tbl-scan-season-seed}
#| label: tbl-scan-season-seed
#| tbl-cap: "BH-significant markers for Seed Index from the season-only scan."

scan_season %>%
  filter(trait == "Seed_index", sig_bh) %>%
  arrange(p.value) %>%
  select(SNP_ID, Chromosome, Pos_cM,
         Estimate = estimate, SE = std.error,
         `$t$` = statistic,
         `$p$-value` = p.value,
         `BH adj. $p$` = p_adj_bh) %>%
  knitr::kable(digits = c(0, 0, 2, 4, 4, 3, 2e-300, 3),
               format.args = list(scientific = TRUE)) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"),
                font_size = 8)
```

## PC-adjusted scan — BH-significant markers

```{r tbl-scan-pc-seed}
#| label: tbl-scan-pc-seed
#| tbl-cap: "BH-significant markers for Seed Index from the PC-adjusted scan."

scan_pc %>%
  filter(trait == "Seed_index", sig_bh) %>%
  arrange(p.value) %>%
  select(SNP_ID, Chromosome, Pos_cM,
         Estimate = estimate, SE = std.error,
         `$t$` = statistic,
         `$p$-value` = p.value,
         `BH adj. $p$` = p_adj_bh) %>%
  knitr::kable(digits = c(0, 0, 2, 4, 4, 3, 2e-300, 3),
               format.args = list(scientific = TRUE)) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"),
                font_size = 8)
```

## Lasso with PC adjustment

```{r tbl-lasso-pc-seed-min}
#| label: tbl-lasso-pc-seed-min
#| tbl-cap: "Markers selected by Lasso + PC for Seed Index under $\\lambda_{\\min}$."

lasso_pc[["Seed_index"]]$min %>%
  left_join(map_df, by = "SNP_ID") %>%
  arrange(desc(abs(coefficient))) %>%
  select(SNP_ID, Chromosome, Pos_cM, `$\\hat{\\beta}$` = coefficient) %>%
  knitr::kable(digits = 4) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8)
```

```{r tbl-lasso-pc-seed-1se}
#| label: tbl-lasso-pc-seed-1se
#| tbl-cap: "Markers selected by Lasso + PC for Seed Index under $\\lambda_{1\\text{se}}$."

lasso_pc[["Seed_index"]]$`1se` %>%
  left_join(map_df, by = "SNP_ID") %>%
  arrange(desc(abs(coefficient))) %>%
  select(SNP_ID, Chromosome, Pos_cM, `$\\hat{\\beta}$` = coefficient) %>%
  knitr::kable(digits = 4) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8)
```

## Lasso with season adjustment only

```{r tbl-lasso-season-seed-min}
#| label: tbl-lasso-season-seed-min
#| tbl-cap: "Markers selected by Lasso season-only for Seed Index under $\\lambda_{\\min}$."

lasso_season[["Seed_index"]]$min %>%
  left_join(map_df, by = "SNP_ID") %>%
  arrange(desc(abs(coefficient))) %>%
  select(SNP_ID, Chromosome, Pos_cM, `$\\hat{\\beta}$` = coefficient) %>%
  knitr::kable(digits = 4) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8)
```

```{r tbl-lasso-season-seed-1se}
#| label: tbl-lasso-season-seed-1se
#| tbl-cap: "Markers selected by Lasso season-only for Seed Index under $\\lambda_{1\\text{se}}$."

lasso_season[["Seed_index"]]$`1se` %>%
  left_join(map_df, by = "SNP_ID") %>%
  arrange(desc(abs(coefficient))) %>%
  select(SNP_ID, Chromosome, Pos_cM, `$\\hat{\\beta}$` = coefficient) %>%
  knitr::kable(digits = 4) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8)
```

{{< pagebreak >}}

# Lint Index

## Season scan — BH-significant markers

```{r tbl-scan-season-lindex}
#| label: tbl-scan-season-lindex
#| tbl-cap: "BH-significant markers for Lint Index from the season-only scan."

scan_season %>%
  filter(trait == "Lint_index", sig_bh) %>%
  arrange(p.value) %>%
  select(SNP_ID, Chromosome, Pos_cM,
         Estimate = estimate, SE = std.error,
         `$t$` = statistic,
         `$p$-value` = p.value,
         `BH adj. $p$` = p_adj_bh) %>%
  knitr::kable(digits = c(0, 0, 2, 4, 4, 3, 2e-300, 3),
               format.args = list(scientific = TRUE)) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8)
```

## PC-adjusted scan — BH-significant markers

```{r tbl-scan-pc-lindex}
#| label: tbl-scan-pc-lindex
#| tbl-cap: "BH-significant markers for Lint Index from the PC-adjusted scan."

scan_pc %>%
  filter(trait == "Lint_index", sig_bh) %>%
  arrange(p.value) %>%
  select(SNP_ID, Chromosome, Pos_cM,
         Estimate = estimate, SE = std.error,
         `$t$` = statistic,
         `$p$-value` = p.value,
         `BH adj. $p$` = p_adj_bh) %>%
  knitr::kable(digits = c(0, 0, 2, 4, 4, 3, 2e-300, 3),
               format.args = list(scientific = TRUE)) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8)
```

## Lasso with PC adjustment

```{r tbl-lasso-pc-lindex-min}
#| label: tbl-lasso-pc-lindex-min
#| tbl-cap: "Markers selected by Lasso + PC for Lint Index under $\\lambda_{\\min}$."

lasso_pc[["Lint_index"]]$min %>%
  left_join(map_df, by = "SNP_ID") %>%
  arrange(desc(abs(coefficient))) %>%
  select(SNP_ID, Chromosome, Pos_cM, `$\\hat{\\beta}$` = coefficient) %>%
  knitr::kable(digits = 4) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8)
```

```{r tbl-lasso-pc-lindex-1se}
#| label: tbl-lasso-pc-lindex-1se
#| tbl-cap: "Markers selected by Lasso + PC for Lint Index under $\\lambda_{1\\text{se}}$."

lasso_pc[["Lint_index"]]$`1se` %>%
  left_join(map_df, by = "SNP_ID") %>%
  arrange(desc(abs(coefficient))) %>%
  select(SNP_ID, Chromosome, Pos_cM, `$\\hat{\\beta}$` = coefficient) %>%
  knitr::kable(digits = 4) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8)
```

## Lasso with season adjustment only

```{r tbl-lasso-season-lindex-min}
#| label: tbl-lasso-season-lindex-min
#| tbl-cap: "Markers selected by Lasso season-only for Lint Index under $\\lambda_{\\min}$."

lasso_season[["Lint_index"]]$min %>%
  left_join(map_df, by = "SNP_ID") %>%
  arrange(desc(abs(coefficient))) %>%
  select(SNP_ID, Chromosome, Pos_cM, `$\\hat{\\beta}$` = coefficient) %>%
  knitr::kable(digits = 4) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8)
```

```{r tbl-lasso-season-lindex-1se}
#| label: tbl-lasso-season-lindex-1se
#| tbl-cap: "Markers selected by Lasso season-only for Lint Index under $\\lambda_{1\\text{se}}$."

lasso_season[["Lint_index"]]$`1se` %>%
  left_join(map_df, by = "SNP_ID") %>%
  arrange(desc(abs(coefficient))) %>%
  select(SNP_ID, Chromosome, Pos_cM, `$\\hat{\\beta}$` = coefficient) %>%
  knitr::kable(digits = 4) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8)
```

{{< pagebreak >}}

# Seed Oil Content

## Season scan — BH-significant markers

```{r tbl-scan-season-oil}
#| label: tbl-scan-season-oil
#| tbl-cap: "BH-significant markers for Seed Oil Content from the season-only scan."

scan_season %>%
  filter(trait == "Seed_oil_content", sig_bh) %>%
  arrange(p.value) %>%
  select(SNP_ID, Chromosome, Pos_cM,
         Estimate = estimate, SE = std.error,
         `$t$` = statistic,
         `$p$-value` = p.value,
         `BH adj. $p$` = p_adj_bh) %>%
  knitr::kable(digits = c(0, 0, 2, 4, 4, 3, 2e-300, 3),
               format.args = list(scientific = TRUE)) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8)
```

## PC-adjusted scan — BH-significant markers

```{r tbl-scan-pc-oil}
#| label: tbl-scan-pc-oil
#| tbl-cap: "BH-significant markers for Seed Oil Content from the PC-adjusted scan."

scan_pc %>%
  filter(trait == "Seed_oil_content", sig_bh) %>%
  arrange(p.value) %>%
  select(SNP_ID, Chromosome, Pos_cM,
         Estimate = estimate, SE = std.error,
         `$t$` = statistic,
         `$p$-value` = p.value,
         `BH adj. $p$` = p_adj_bh) %>%
  knitr::kable(digits = c(0, 0, 2, 4, 4, 3, 2e-300, 3),
               format.args = list(scientific = TRUE)) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8)
```

## Lasso with PC adjustment

```{r tbl-lasso-pc-oil-min}
#| label: tbl-lasso-pc-oil-min
#| tbl-cap: "Markers selected by Lasso + PC for Seed Oil Content under $\\lambda_{\\min}$."

lasso_pc[["Seed_oil_content"]]$min %>%
  left_join(map_df, by = "SNP_ID") %>%
  arrange(desc(abs(coefficient))) %>%
  select(SNP_ID, Chromosome, Pos_cM, `$\\hat{\\beta}$` = coefficient) %>%
  knitr::kable(digits = 4) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8)
```

```{r tbl-lasso-pc-oil-1se}
#| label: tbl-lasso-pc-oil-1se
#| tbl-cap: "Markers selected by Lasso + PC for Seed Oil Content under $\\lambda_{1\\text{se}}$."

lasso_pc[["Seed_oil_content"]]$`1se` %>%
  left_join(map_df, by = "SNP_ID") %>%
  arrange(desc(abs(coefficient))) %>%
  select(SNP_ID, Chromosome, Pos_cM, `$\\hat{\\beta}$` = coefficient) %>%
  knitr::kable(digits = 4) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8)
```

## Lasso with season adjustment only

```{r tbl-lasso-season-oil-min}
#| label: tbl-lasso-season-oil-min
#| tbl-cap: "Markers selected by Lasso season-only for Seed Oil Content under $\\lambda_{\\min}$."

lasso_season[["Seed_oil_content"]]$min %>%
  left_join(map_df, by = "SNP_ID") %>%
  arrange(desc(abs(coefficient))) %>%
  select(SNP_ID, Chromosome, Pos_cM, `$\\hat{\\beta}$` = coefficient) %>%
  knitr::kable(digits = 4) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8)
```

```{r tbl-lasso-season-oil-1se}
#| label: tbl-lasso-season-oil-1se
#| tbl-cap: "Markers selected by Lasso season-only for Seed Oil Content under $\\lambda_{1\\text{se}}$."

lasso_season[["Seed_oil_content"]]$`1se` %>%
  left_join(map_df, by = "SNP_ID") %>%
  arrange(desc(abs(coefficient))) %>%
  select(SNP_ID, Chromosome, Pos_cM, `$\\hat{\\beta}$` = coefficient) %>%
  knitr::kable(digits = 4) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8)
```

{{< pagebreak >}}

# Cross-Trait Summary

## Markers appearing in multiple traits (scan, BH-significant)

```{r tbl-shared-scan-markers}
#| label: tbl-shared-scan-markers
#| tbl-cap: "Markers that are BH-significant in two or more traits under the PC-adjusted scan, with their estimated effects per trait."

shared_scan <- scan_pc %>%
  filter(sig_bh) %>%
  group_by(SNP_ID) %>%
  filter(n_distinct(trait) >= 2) %>%
  ungroup() %>%
  select(SNP_ID, trait, Chromosome, Pos_cM, estimate, p.value, p_adj_bh) %>%
  mutate(trait = gsub("_", " ", trait)) %>%
  arrange(SNP_ID, trait)

if (nrow(shared_scan) == 0) {
  cat("No markers are BH-significant in two or more traits simultaneously.")
} else {
  shared_scan %>%
    rename(Trait = trait, Estimate = estimate,
           `$p$` = p.value, `BH adj. $p$` = p_adj_bh) %>%
    knitr::kable(digits = 4) %>%
    kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8)
}
```

## Markers appearing in multiple traits (lasso, $\lambda_{1\text{se}}$)

```{r tbl-shared-lasso-markers}
#| label: tbl-shared-lasso-markers
#| tbl-cap: "Markers selected by Lasso + PC ($\\lambda_{1\\text{se}}$) in two or more traits."

lasso_1se_long <- bind_rows(lapply(trait_names, function(tr) {
  lasso_pc[[tr]]$`1se` %>%
    mutate(trait = gsub("_", " ", tr))
}))

shared_lasso <- lasso_1se_long %>%
  group_by(SNP_ID) %>%
  filter(n_distinct(trait) >= 2) %>%
  ungroup() %>%
  left_join(map_df, by = "SNP_ID") %>%
  select(SNP_ID, Chromosome, Pos_cM, trait, coefficient) %>%
  arrange(SNP_ID, trait)

if (nrow(shared_lasso) == 0) {
  cat("No markers are selected by lasso in two or more traits simultaneously.")
} else {
  shared_lasso %>%
    rename(Trait = trait, `$\\hat{\\beta}$` = coefficient) %>%
    knitr::kable(digits = 4) %>%
    kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8)
}
```

## Overall selection count summary

```{r tbl-overall-count-summary}
#| label: tbl-overall-count-summary
#| tbl-cap: "Total number of markers selected by each method and trait combination. Scan results use the BH threshold; lasso results are shown for both $\\lambda$ rules."

count_summary <- bind_rows(
  lapply(trait_names, function(tr) {
    tibble(
      Trait                    = gsub("_", " ", tr),
      `Season scan (BH)`       = sum(scan_season$trait == tr & scan_season$sig_bh,
                                     na.rm = TRUE),
      `PC scan (BH)`           = sum(scan_pc$trait     == tr & scan_pc$sig_bh,
                                     na.rm = TRUE),
      `Lasso+PC (min)`         = nrow(lasso_pc[[tr]]$min),
      `Lasso+PC (1se)`         = nrow(lasso_pc[[tr]]$`1se`),
      `Lasso season (min)`     = nrow(lasso_season[[tr]]$min),
      `Lasso season (1se)`     = nrow(lasso_season[[tr]]$`1se`)
    )
  })
)

knitr::kable(count_summary) %>%
  kable_styling(latex_options = "hold_position", font_size = 9)
```
