---
title: "From Classical Single-Marker QTL Mapping to Multi-Marker Shrinkage: Analysis and Simulation Using Cotton MAGIC Genotypes"
author: "Shiyan Miao"
date: today
format:
  # revealjs: default
  pdf:
    documentclass: article
    toc: true
    number-sections: true
    colorlinks: true
    geometry:
      - margin=1in
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

## Required packages

```{r}
library(tidyverse)
library(data.table)
library(glmnet)
library(broom)
library(janitor)
library(patchwork)
library(Matrix)
library(knitr)
library(scales)
library(irlba)
```

## File paths

```{r}
geno_file <- "Genotype_data.txt"
map_file  <- "Genetic_map.txt"
pheno_file <- "Phenotype_data.txt"
```

## Import data

```{r}
geno_raw <- fread(geno_file, header = FALSE)
map_df <- fread(map_file)
pheno_raw <- fread(pheno_file)
```

## Inspect dimensions

```{r}
dim(geno_raw)
dim(map_df)
dim(pheno_raw)

head(map_df)
head(pheno_raw)
```

This block establishes the analytical dataset. The report confirms a **256 × 6523** genotype matrix, a **6523 × 3** marker map, and a **768 × 4** phenotype file. On that basis, it infers that the phenotype file represents **256 lines observed across 3 stacked seasons/environments**, reconstructs `Line` and `Season`, and merges phenotype with genotype into `analysis_df` with **768 rows and 6530 columns**.

## Add marker names to the genotype matrix

The genotype file has no column names. We use the SNP identifiers from the genetic map.

```{r}
stopifnot(ncol(geno_raw) == nrow(map_df))

geno_df <- as.data.frame(geno_raw)
colnames(geno_df) <- map_df$SNP_ID
geno_df$Line <- 1:nrow(geno_df)

geno_df[1:5, 1:8]
```

## Construct line and season identifiers for phenotype data

The phenotype file contains 768 rows, while the genotype matrix contains 256 lines. This strongly suggests that phenotype observations correspond to three stacked seasons or environments for the same 256 lines.

```{r}
n_lines <- nrow(geno_df)
n_pheno <- nrow(pheno_raw)

n_pheno / n_lines
```

If the phenotype observations are indeed stacked by season, then each block of 256 rows corresponds to one season.

```{r}
pheno_df <- pheno_raw %>%
  mutate(
    RowID = row_number(),
    Season = case_when(
      RowID <= n_lines ~ "Season_1",
      RowID <= 2 * n_lines ~ "Season_2",
      TRUE ~ "Season_3"
    ),
    Line = rep(1:n_lines, times = n_pheno / n_lines)
  ) %>%
  relocate(Line, Season, RowID)

head(pheno_df)
tail(pheno_df)
```

## Merge phenotype and genotype

```{r}
analysis_df <- pheno_df %>%
  left_join(geno_df, by = "Line")

dim(analysis_df)
```

# Descriptive Analysis

## Trait names

```{r}
trait_names <- c(
  "Lint_percentage",
  "Seed_index",
  "Lint_index",
  "Seed_oil_content"
)
```

## Summary statistics

```{r}
trait_summary <- analysis_df %>%
  summarise(across(all_of(trait_names), list(
    mean = ~mean(.x, na.rm = TRUE),
    sd = ~sd(.x, na.rm = TRUE),
    min = ~min(.x, na.rm = TRUE),
    max = ~max(.x, na.rm = TRUE)
  )))

trait_summary
```

## Season-specific summaries

```{r}
season_summary <- analysis_df %>%
  group_by(Season) %>%
  summarise(across(all_of(trait_names), list(
    mean = ~mean(.x, na.rm = TRUE),
    sd = ~sd(.x, na.rm = TRUE)
  )), .groups = "drop")

season_summary
```

## Trait distributions

```{r}
plot_trait_density <- function(df, trait) {
  ggplot(df, aes(x = .data[[trait]], fill = Season)) +
    geom_density(alpha = 0.4) +
    labs(
      title = paste("Distribution of", trait, "by season"),
      x = trait,
      y = "Density"
    ) +
    theme_minimal()
}

plots_density <- lapply(trait_names, function(tr) plot_trait_density(analysis_df, tr))

plots_density[[1]]
plots_density[[2]]
plots_density[[3]]
plots_density[[4]]
```

## Trait boxplots by season

```{r}
plot_trait_box <- function(df, trait) {
  ggplot(df, aes(x = Season, y = .data[[trait]], fill = Season)) +
    geom_boxplot(alpha = 0.8) +
    labs(
      title = paste("Boxplot of", trait, "by season"),
      x = "Season",
      y = trait
    ) +
    theme_minimal()
}

plots_box <- lapply(trait_names, function(tr) plot_trait_box(analysis_df, tr))

plots_box[[1]]
plots_box[[2]]
plots_box[[3]]
plots_box[[4]]
```

The report focuses on four traits: `Lint_percentage`, `Seed_index`, `Lint_index`, and `Seed_oil_content`. Across seasons, the report shows strong mean shifts: `Lint_percentage` rises from **39.5** in Season_1 to **43.9** in Seasons_2 and \_3, while `Seed_index` declines from **9.58** to **8.15** and then **7.73**.

These figures show that season effects are not driven by a handful of extreme observations; they reflect **systematic distributional shifts**. They also show that substantial within-season variation remains, which is exactly what makes QTL analysis feasible.

However, the plots do not formally diagnose whether residual variance is homogeneous across seasons, nor do they account for repeated observations of the same line across seasons. So they are strong descriptive diagnostics, but not a substitute for full model checking.

## Correlation among traits

```{r}
trait_cor <- analysis_df %>%
  select(all_of(trait_names)) %>%
  cor(use = "pairwise.complete.obs")

trait_cor
```

The correlation matrix shows a very strong negative association between `Lint_percentage` and `Seed_index` (**-0.778**), a moderate positive association between `Lint_percentage` and `Lint_index` (**0.354**), a weaker negative association between `Seed_index` and `Seed_oil_content` (**-0.277**), and near-zero correlation between `Seed_oil_content` and `Lint_percentage` (**0.057**).

This is an important structural result. The strong `Lint_percentage`–`Seed_index` trade-off suggests that hits shared across those traits may reflect a common underlying axis rather than independent mechanisms. By contrast, `Seed_oil_content` appears much more weakly connected to the others, implying a comparatively distinct genetic dimension. In substantive terms, the correlation matrix already suggests a system with at least one major **lint–seed trade-off axis** and a more separate **oil-content axis**.

```{r}
head(geno_raw[, 1:min(8, ncol(geno_raw))])
head(map_df)
head(pheno_raw)
```

The purpose of this step is to confirm that the imported files match the expected Week-3 data structure. In the previous report, this dataset consisted of a 256-line genotype matrix, a 6523-marker map, and a 768-row phenotype file. If the imported dimensions are broadly consistent with this structure, then the current analysis can proceed on the same empirical basis as the earlier report.

```{r}
stopifnot(ncol(geno_raw) == nrow(map_df))

geno_df <- as.data.frame(geno_raw)
colnames(geno_df) <- map_df$SNP_ID
geno_df$Line <- 1:nrow(geno_df)

knitr::kable(head(map_df))
```

The raw genotype file does not contain marker names, so it cannot be meaningfully interpreted or used for PCA in its imported form. This step therefore aligns the genotype matrix with the marker map by assigning the `SNP_ID` values from the map file as column names. It also introduces a `Line` variable, which is necessary for matching genotype rows to reconstructed phenotype line identities.

```{r}
n_lines <- nrow(geno_df)
n_pheno <- nrow(pheno_raw)

pheno_df <- pheno_raw %>%
  mutate(
    RowID = row_number(),
    Season = case_when(
      RowID <= n_lines ~ "Season_1",
      RowID <= 2 * n_lines ~ "Season_2",
      TRUE ~ "Season_3"
    ),
    Line = rep(1:n_lines, times = n_pheno / n_lines)
  ) %>%
  relocate(Line, Season, RowID)

head(pheno_df)
```

This step reconstructs the phenotype structure by assigning both a `Line` identifier and a `Season` label, thereby creating a bridge between the phenotype and genotype data.

```{r}
analysis_df <- pheno_df %>%
  left_join(geno_df, by = "Line")

merge_tbl <- tibble(
  object = c("analysis_df"),
  nrow = nrow(analysis_df),
  ncol = ncol(analysis_df)
)

knitr::kable(merge_tbl)
```

This step merges phenotype and genotype information by line. Although PCA itself only requires the genotype matrix, integrating phenotype into the broader analysis object is useful for two reasons: it allows later interpretation of the PCs in relation to the traits, and it enables phenotype-based colouring or annotation in PCA visualisations.

```{r}
X <- geno_df %>%
  select(-Line) %>%
  as.matrix()

mode(X) <- "numeric"

for (j in seq_len(ncol(X))) {
  miss_idx <- is.na(X[, j])
  if (any(miss_idx)) {
    X[miss_idx, j] <- mean(X[, j], na.rm = TRUE)
  }
}
```

PCA is applied only to the genotype matrix, so this section extracts the marker matrix from `geno_df`, removes the `Line` identifier, and converts the result into a numeric matrix `X`. Any missing values are mean-imputed on a marker-by-marker basis for PCA diagnostics. This imputation is used only to enable dimensional reduction and does not imply that the same strategy must be used in later modelling.

```{r}
set.seed(123)

pca_fit <- prcomp_irlba(
  x = scale(X, center = TRUE, scale = TRUE),
  n = 20,
  center = FALSE,
  scale. = FALSE
)

var_explained <- (pca_fit$sdev^2) / sum(pca_fit$sdev^2)
cum_var_explained <- cumsum(var_explained)

pca_var_df <- tibble(
  PC = 1:length(var_explained),
  Variance_Explained = var_explained,
  Cumulative = cum_var_explained
)
```

PCA is now applied to the genotype matrix using `prcomp_irlba()`, which is more efficient than standard PCA for larger matrices. The purpose here is not direct prediction, but structural diagnosis: to determine whether the genotype matrix contains clear lower-dimensional variation patterns, and to assess whether these patterns may matter for later multiple-marker analysis.

```{r}
ggplot(pca_var_df, aes(x = PC, y = Variance_Explained)) +
  geom_col() +
  labs(
    title = "Scree Plot for Genotype PCA",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  theme_minimal(base_size = 14)
```

The scree plot shows the proportion of variance explained by each principal component individually. If the first few PCs explain a substantial share of the variance, this suggests that the genotype matrix contains strong low-dimensional structure. If variance is spread thinly across many components, then the data are better characterised as highly dispersed and structurally complex.

The scree plot indicates that no single principal component dominates the genotype variation; even the leading PCs explain only modest proportions of variance. The cumulative variance plot shows that multiple PCs are needed to capture a substantial share of total structure. That implies a **diffuse, multi-dimensional genotype structure** rather than a single overwhelming stratification axis.

```{r}
ggplot(pca_var_df, aes(x = PC, y = Cumulative)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Cumulative Variance Explained",
    x = "Principal Component",
    y = "Cumulative Proportion"
  ) +
  theme_minimal(base_size = 14)
```

The cumulative variance plot shows how much total genotype variation is captured as more principal components are included. This is useful for evaluating whether the first 5, 10, or 20 PCs provide an adequate summary of the overall marker structure.

```{r}
pca_scores <- as_tibble(pca_fit$x[, 1:5]) %>%
  mutate(Line = 1:n())

trait_line_mean <- pheno_df %>%
  group_by(Line) %>%
  summarise(
    Lint_percentage = mean(Lint_percentage, na.rm = TRUE),
    Seed_index = mean(Seed_index, na.rm = TRUE),
    Lint_index = mean(Lint_index, na.rm = TRUE),
    Seed_oil_content = mean(Seed_oil_content, na.rm = TRUE),
    .groups = "drop"
  )

pca_scores <- pca_scores %>%
  left_join(trait_line_mean, by = "Line")
```

This section extracts the PCA scores for each line on the first few principal components. It then merges in line-level mean trait values, averaged across seasons, in order to support later comparison between the genetic PCs and phenotype patterns.

```{r}
ggplot(pca_scores, aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.7) +
  labs(
    title = "Genotype PCA: PC1 vs PC2",
    x = "PC1",
    y = "PC2"
  ) +
  theme_minimal(base_size = 14)
```

PC1 and PC2 typically capture the dominant axes of variation in the genotype matrix. This plot is therefore used to inspect whether the 256 lines show visible clustering, gradients, or broad separation in the first two principal components.

```{r}
ggplot(pca_scores, aes(x = PC1, y = PC3)) +
  geom_point(alpha = 0.7) +
  labs(
    title = "Genotype PCA: PC1 vs PC3",
    x = "PC1",
    y = "PC3"
  ) +
  theme_minimal(base_size = 14)
```

It is often useful to inspect PC1 against PC3 as well, because PC3 may capture variation that is not visible in PC2. This provides a broader picture of whether there are additional structural patterns in the genotype matrix.

```{r}
ggplot(pca_scores, aes(x = PC1, y = PC2, colour = Lint_percentage)) +
  geom_point(size = 2, alpha = 0.8) +
  labs(
    title = "PC1 vs PC2 coloured by average lint percentage",
    x = "PC1",
    y = "PC2",
    colour = "Lint %"
  ) +
  theme_minimal(base_size = 14)
```

This figure colours the PCA plot by the line-level mean of lint percentage. The purpose is not to claim that the PCs directly predict the trait, but to visually inspect whether phenotype variation appears to align with the major genomic axes.

```{r}
pc_trait_df <- pca_scores %>%
  select(Line, PC1, PC2, PC3, PC4, PC5, Lint_percentage, Seed_index, Lint_index, Seed_oil_content)

cor_mat <- pc_trait_df %>%
  select(-Line) %>%
  cor(use = "pairwise.complete.obs")

cor_df <- as.data.frame(cor_mat) %>%
  rownames_to_column("Var1") %>%
  pivot_longer(-Var1, names_to = "Var2", values_to = "Correlation") %>%
  filter(
    Var1 %in% c("PC1", "PC2", "PC3", "PC4", "PC5"),
    Var2 %in% c("Lint_percentage", "Seed_index", "Lint_index", "Seed_oil_content")
  )

ggplot(cor_df, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile() +
  geom_text(aes(label = round(Correlation, 2)), size = 5) +
  scale_fill_gradient2(low = "darkblue", mid = "white", high = "darkred", midpoint = 0) +
  labs(
    title = "Correlation between top PCs and averaged traits",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 14)
```

The PC1–PC2 and PC1–PC3 scatterplots show that the 256 lines are not randomly dispersed; there is visible broad separation and gradient-like structure, especially along PC1, but not sharply isolated clusters. The colour-coded `Lint_percentage` PCA plot does not show a clean phenotypic partition along the first two genotype axes. The heatmap supports that interpretation: PC–trait correlations are generally modest, with the strongest values being **PC4–Lint_index = 0.32**, **PC5–Lint_index = 0.21**, and **PC5–Seed_oil_content = 0.19**.

**Population structure is present and worth controlling, but it does not by itself explain the phenotype patterns.** This justifies the use of top PCs as broad structural covariates rather than as primary trait predictors.

The limitations are also clear. Missing genotypes were mean-imputed marker by marker for PCA, which is acceptable for exploratory dimensional diagnosis but may change tail variation. The choice to include only **PC1–PC5** is sensible, but still heuristic. Most importantly, because the scatterplots suggest continuous relatedness rather than neat discrete clusters, PC adjustment alone may be insufficient; a kinship-aware mixed modelling strategy would be stronger.

The baseline single-marker model is:

$$
y_{is} = \alpha + \beta_j G_{ij} + \gamma_s \,\mathrm{Season}_s + \varepsilon_{is}
$$

where $y_{is}$ is the phenotype for line $i$ in season $s$, $G_{ij}$ is the genotype of marker $j$ for line $i$, and Season is included as a control.

The PC-adjusted single-marker model is:

$$
y_{is} = \alpha + \beta_j G_{ij} + \gamma_s \,\mathrm{Season}_s + \sum_{k=1}^{5} \delta_k PC_{ik} + \varepsilon_{is}
$$

where the first five genotype principal components are included as additional covariates to account for broad population structure.

```{r}
# Merge top 5 genotype PCs back into analysis_df

pc_line_df <- pca_scores %>%
  dplyr::select(Line, PC1, PC2, PC3, PC4, PC5)

analysis_df <- analysis_df %>%
  dplyr::left_join(pc_line_df, by = "Line")

analysis_df$Season <- factor(analysis_df$Season)
marker_names <- colnames(geno_df)[colnames(geno_df) != "Line"]
```

```{r}
# Prepare marker matrix
prep_marker_matrix <- function(df, marker_names) {
  X <- as.matrix(df[, marker_names, drop = FALSE])
  mode(X) <- "numeric"
  
  if (anyNA(X)) {
    col_means <- colMeans(X, na.rm = TRUE)
    idx <- which(is.na(X), arr.ind = TRUE)
    X[idx] <- col_means[idx[, 2]]
  }
  
  X
}
```

```{r}
# Control Seasons and PCs
fast_single_marker_scan_cov <- function(df, trait, marker_names, covariates = "Season") {
  df <- as.data.frame(df)
  
  needed_cols <- unique(c(trait, covariates, marker_names))
  df <- df[, needed_cols, drop = FALSE]
  
  keep <- complete.cases(df[, c(trait, covariates), drop = FALSE])
  df <- df[keep, , drop = FALSE]
  
  y <- df[[trait]]
  X <- prep_marker_matrix(df, marker_names)
  
  cov_formula <- as.formula(
    paste("~", paste(covariates, collapse = " + "))
  )
  C <- model.matrix(cov_formula, data = df)
  qrC <- qr(C)
  
  # Residualise phenotype and marker matrix on covariates
  y_res <- qr.resid(qrC, y)
  X_res <- qr.resid(qrC, X)
  
  Sxx <- colSums(X_res^2)
  Sxy <- as.vector(crossprod(X_res, y_res))
  Syy <- sum(y_res^2)
  
  df_resid <- nrow(df) - qrC$rank - 1
  valid <- Sxx > 1e-12
  
  beta <- rep(NA_real_, length(marker_names))
  rss  <- rep(NA_real_, length(marker_names))
  se   <- rep(NA_real_, length(marker_names))
  tval <- rep(NA_real_, length(marker_names))
  pval <- rep(NA_real_, length(marker_names))
  
  beta[valid] <- Sxy[valid] / Sxx[valid]
  rss[valid]  <- pmax(Syy - (Sxy[valid]^2 / Sxx[valid]), 0)
  se[valid]   <- sqrt((rss[valid] / df_resid) / Sxx[valid])
  tval[valid] <- beta[valid] / se[valid]
  pval[valid] <- 2 * pt(abs(tval[valid]), df = df_resid, lower.tail = FALSE)
  
  tibble::tibble(
    trait = trait,
    SNP_ID = marker_names,
    estimate = beta,
    std.error = se,
    statistic = tval,
    p.value = pval,
    covariates = paste(covariates, collapse = " + "),
    df_resid = df_resid
  )
}
```

```{r}
# Covariate sets

covars_season_only <- c("Season")
covars_season_pc   <- c("Season", "PC1", "PC2", "PC3", "PC4", "PC5")

# Run scans for all traits 

scan_season_only <- lapply(trait_names, function(tr) {
  fast_single_marker_scan_cov(
    df = analysis_df,
    trait = tr,
    marker_names = marker_names,
    covariates = covars_season_only
  )
}) %>%
  dplyr::bind_rows() %>%
  dplyr::left_join(map_df, by = "SNP_ID") %>%
  dplyr::mutate(log10p = -log10(p.value))

scan_season_pc <- lapply(trait_names, function(tr) {
  fast_single_marker_scan_cov(
    df = analysis_df,
    trait = tr,
    marker_names = marker_names,
    covariates = covars_season_pc
  )
}) %>%
  dplyr::bind_rows() %>%
  dplyr::left_join(map_df, by = "SNP_ID") %>%
  dplyr::mutate(log10p = -log10(p.value))
```

```{r}
# Multiple-testing correction-
add_multiple_testing <- function(scan_df, alpha = 0.05) {
  scan_df %>%
    dplyr::group_by(trait, covariates) %>%
    dplyr::arrange(p.value, .by_group = TRUE) %>%
    dplyr::mutate(
      p_adj_bh   = p.adjust(p.value, method = "BH"),
      p_adj_bonf = p.adjust(p.value, method = "bonferroni"),
      sig_raw    = p.value < alpha,
      sig_bh     = p_adj_bh < alpha,
      sig_bonf   = p_adj_bonf < alpha
    ) %>%
    dplyr::ungroup()
}

scan_season_only_adj <- add_multiple_testing(scan_season_only)
scan_season_pc_adj   <- add_multiple_testing(scan_season_pc)
```

```{r}
sig_summary <- dplyr::bind_rows(
  scan_season_only_adj %>% dplyr::mutate(model = "Season only"),
  scan_season_pc_adj %>% dplyr::mutate(model = "Season + PC1-5")
) %>%
  dplyr::group_by(model, trait) %>%
  dplyr::summarise(
    n_raw  = sum(sig_raw, na.rm = TRUE),
    n_bh   = sum(sig_bh, na.rm = TRUE),
    n_bonf = sum(sig_bonf, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(sig_summary, digits = 3)
```

```{r}
trait_to_plot <- "Lint_percentage"

top_markers_compare <- scan_season_only_adj %>%
  dplyr::filter(trait == trait_to_plot) %>%
  dplyr::arrange(p.value) %>%
  dplyr::slice_head(n = 20) %>%
  dplyr::select(
    SNP_ID,
    estimate_season = estimate,
    p_season = p.value,
    log10p_season = log10p,
    sig_bh_season = sig_bh,
    sig_bonf_season = sig_bonf
  ) %>%
  dplyr::left_join(
    scan_season_pc_adj %>%
      dplyr::filter(trait == trait_to_plot) %>%
      dplyr::select(
        SNP_ID,
        estimate_pc = estimate,
        p_pc = p.value,
        log10p_pc = log10p,
        sig_bh_pc = sig_bh,
        sig_bonf_pc = sig_bonf
      ),
    by = "SNP_ID"
  ) %>%
  dplyr::mutate(
    abs_est_change = abs(estimate_pc - estimate_season),
    p_ratio = p_pc / p_season
  )

knitr::kable(top_markers_compare, digits = 4)
```

```{r}
select_marker_set <- function(scan_df, trait_name, rule = c("bh", "bonf", "top"), top_n = 50) {
  rule <- match.arg(rule)
  
  df <- scan_df %>%
    dplyr::filter(trait == trait_name) %>%
    dplyr::arrange(p.value)
  
  if (rule == "bh") {
    hits <- df %>% dplyr::filter(sig_bh) %>% dplyr::pull(SNP_ID)
    return(unique(hits))
  }
  
  if (rule == "bonf") {
    hits <- df %>% dplyr::filter(sig_bonf) %>% dplyr::pull(SNP_ID)
    return(unique(hits))
  }
  
  df %>% dplyr::slice_head(n = top_n) %>% dplyr::pull(SNP_ID) %>% unique()
}
```

```{r}
bh_season <- select_marker_set(scan_season_only_adj, trait_to_plot, rule = "bh")
bh_pc     <- select_marker_set(scan_season_pc_adj, trait_to_plot, rule = "bh")

overlap_summary_bh <- tibble::tibble(
  trait = trait_to_plot,
  rule = "BH-FDR",
  n_season = length(bh_season),
  n_pc = length(bh_pc),
  n_overlap = length(intersect(bh_season, bh_pc)),
  jaccard = ifelse(
    length(union(bh_season, bh_pc)) == 0,
    NA_real_,
    length(intersect(bh_season, bh_pc)) / length(union(bh_season, bh_pc))
  )
)

knitr::kable(overlap_summary_bh, digits = 3)
```

$$ J(A, B) = \frac{|A \cap B|}{|A \cup B|}$$

```{r}
bh_season <- select_marker_set(scan_season_only_adj, trait_to_plot, rule = "bh")
bh_pc     <- select_marker_set(scan_season_pc_adj, trait_to_plot, rule = "bh")

overlap_summary_bh <- tibble::tibble(
  trait = trait_to_plot,
  rule = "BH-FDR",
  n_season = length(bh_season),
  n_pc = length(bh_pc),
  n_overlap = length(intersect(bh_season, bh_pc)),
  jaccard = ifelse(
    length(union(bh_season, bh_pc)) == 0,
    NA_real_,
    length(intersect(bh_season, bh_pc)) / length(union(bh_season, bh_pc))
  )
)

knitr::kable(overlap_summary_bh, digits = 3)
```

```{r}
sig_summary_long <- sig_summary %>%
  tidyr::pivot_longer(
    cols = c(n_raw, n_bh, n_bonf),
    names_to = "threshold",
    values_to = "n_significant"
  )

ggplot2::ggplot(sig_summary_long, ggplot2::aes(x = threshold, y = n_significant, fill = model)) +
  ggplot2::geom_col(position = "dodge") +
  ggplot2::facet_wrap(~ trait, scales = "free_y") +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::labs(
    title = "Number of significant markers across model specifications",
    x = NULL,
    y = "Number of significant markers"
  )
```

Using `Lint_percentage` as the example trait, the report compares the top markers under both model specifications. Many top hits, including markers such as `Marker-4017`, `Marker-4013`, and `Marker-4000`, retain very similar effect sizes across models. At the same time, some markers shift substantially in significance. Under BH, the overlap between the two significant-marker sets is **1090**, with a **Jaccard index of 0.602**.

The interpretation is that **PC adjustment changes the boundary of the significance map more than it overturns the strongest signals outright**. Strong leading hits are moderately robust, but the broader set of candidate markers is model-sensitive. A Jaccard index of 0.602 indicates meaningful overlap, but certainly not equivalence.

Accordingly, the most credible candidate regions are those that remain highly ranked across both model specifications and continue to appear under shrinkage selection.

```{r}
compare_pvals <- scan_season_only_adj %>%
  dplyr::select(trait, SNP_ID, log10p_season = log10p, est_season = estimate) %>%
  dplyr::left_join(
    scan_season_pc_adj %>%
      dplyr::select(trait, SNP_ID, log10p_pc = log10p, est_pc = estimate),
    by = c("trait", "SNP_ID")
  )
```

```{r}
# ---- Compare season-only vs season+PC marker statistics ----

compare_scan_df <- scan_season_only_adj %>%
  dplyr::select(
    trait, SNP_ID,
    p_season = p.value,
    log10p_season = log10p,
    estimate_season = estimate,
    p_adj_bh_season = p_adj_bh,
    p_adj_bonf_season = p_adj_bonf
  ) %>%
  dplyr::left_join(
    scan_season_pc_adj %>%
      dplyr::select(
        trait, SNP_ID,
        p_pc = p.value,
        log10p_pc = log10p,
        estimate_pc = estimate,
        p_adj_bh_pc = p_adj_bh,
        p_adj_bonf_pc = p_adj_bonf
      ),
    by = c("trait", "SNP_ID")
  )
```

```{r}
trait_to_plot <- "Lint_percentage"

ggplot2::ggplot(
  compare_scan_df %>% dplyr::filter(trait == trait_to_plot),
  ggplot2::aes(x = log10p_season, y = log10p_pc)
) +
  ggplot2::geom_point(alpha = 0.35, size = 0.8) +
  ggplot2::geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::labs(
    title = paste("Season only vs Season + PC1-5:", trait_to_plot),
    x = expression(-log[10](p)~"(Season only)"),
    y = expression(-log[10](p)~"(Season + PC1-5)")
  )
```

```{r}
ggplot2::ggplot(
  compare_scan_df %>% dplyr::filter(trait == trait_to_plot),
  ggplot2::aes(x = estimate_season, y = estimate_pc)
) +
  ggplot2::geom_point(alpha = 0.35, size = 0.8) +
  ggplot2::geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::labs(
    title = paste("Marker effect estimates:", trait_to_plot),
    x = "Estimate (Season only)",
    y = "Estimate (Season + PC1-5)"
  )
```

```{r}
scan_diff_summary <- compare_scan_df %>%
  dplyr::group_by(trait) %>%
  dplyr::summarise(
    cor_log10p = cor(log10p_season, log10p_pc, use = "complete.obs"),
    cor_estimate = cor(estimate_season, estimate_pc, use = "complete.obs"),
    mean_abs_est_diff = mean(abs(estimate_pc - estimate_season), na.rm = TRUE),
    n_bh_season = sum(p_adj_bh_season < 0.05, na.rm = TRUE),
    n_bh_pc = sum(p_adj_bh_pc < 0.05, na.rm = TRUE),
    n_bonf_season = sum(p_adj_bonf_season < 0.05, na.rm = TRUE),
    n_bonf_pc = sum(p_adj_bonf_pc < 0.05, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(scan_diff_summary, digits = 3)
```

```{r}
trait_to_check <- "Lint_percentage"

top20_markers <- scan_season_only_adj %>%
  dplyr::filter(trait == trait_to_check) %>%
  dplyr::arrange(p.value) %>%
  dplyr::slice_head(n = 20) %>%
  dplyr::pull(SNP_ID)
```

```{r}
# ---- Create season-residualised traits ----

make_season_residuals <- function(df, trait_names) {
  df <- as.data.frame(df)
  
  if (!"Season" %in% names(df)) {
    stop("Column 'Season' not found in analysis_df.")
  }
  
  for (tr in trait_names) {
    if (!tr %in% names(df)) {
      stop(paste("Trait not found in analysis_df:", tr))
    }
    
    tmp <- data.frame(
      .y = df[[tr]],
      .season = df[["Season"]]
    )
    
    keep <- stats::complete.cases(tmp)
    resid_vec <- rep(NA_real_, nrow(df))
    
    fit <- stats::lm(.y ~ .season, data = tmp[keep, , drop = FALSE])
    resid_vec[keep] <- stats::resid(fit)
    
    df[[paste0(tr, "_season_resid")]] <- resid_vec
  }
  
  df
}

analysis_df_resid <- make_season_residuals(analysis_df, trait_names)

cat("Created analysis_df_resid\n")
print(dim(analysis_df_resid))
print(colnames(analysis_df_resid)[grepl("_season_resid$", colnames(analysis_df_resid))])
```

```{r}
# ---- Exact LM on season-residualised phenotype ----

run_exact_resid_lm <- function(df, resid_trait, marker_names) {
  df <- as.data.frame(df)
  
  if (!resid_trait %in% names(df)) {
    stop(paste("Residual trait not found in df:", resid_trait))
  }
  
  out <- lapply(marker_names, function(mk) {
    if (!mk %in% names(df)) {
      stop(paste("Marker not found in df:", mk))
    }
    
    tmp <- data.frame(
      .y = df[[resid_trait]],
      .marker = df[[mk]]
    )
    
    tmp <- tmp[stats::complete.cases(tmp), , drop = FALSE]
    
    if (nrow(tmp) < 3 || stats::var(tmp$.marker) == 0) {
      return(tibble::tibble(
        resid_trait = resid_trait,
        SNP_ID = mk,
        estimate = NA_real_,
        std.error = NA_real_,
        statistic = NA_real_,
        p.value = NA_real_
      ))
    }
    
    fit <- stats::lm(.y ~ .marker, data = tmp)
    co <- broom::tidy(fit)
    co <- co[co$term == ".marker", , drop = FALSE]
    
    tibble::tibble(
      resid_trait = resid_trait,
      SNP_ID = mk,
      estimate = co$estimate,
      std.error = co$std.error,
      statistic = co$statistic,
      p.value = co$p.value
    )
  })
  
  dplyr::bind_rows(out)
}
```

```{r}
exact_resid_lm_results <- run_exact_resid_lm(
  df = analysis_df_resid,
  resid_trait = paste0(trait_to_check, "_season_resid"),
  marker_names = top20_markers
)

knitr::kable(exact_resid_lm_results, digits = 4)
```

```{r}
compare_resid_vs_scan <- scan_season_only_adj %>%
  dplyr::filter(trait == trait_to_check, SNP_ID %in% top20_markers) %>%
  dplyr::select(
    SNP_ID,
    estimate_scan = estimate,
    p_scan = p.value,
    log10p_scan = log10p
  ) %>%
  dplyr::left_join(
    exact_resid_lm_results %>%
      dplyr::select(
        SNP_ID,
        estimate_resid_lm = estimate,
        p_resid_lm = p.value
      ),
    by = "SNP_ID"
  )

knitr::kable(compare_resid_vs_scan, digits = 4)
```

The report residualises `Lint_percentage` on season and reruns exact linear models for the top 20 markers. The estimates and p-values closely reproduce the results from the fast scan.

This is an **implementation validation**, not a new biological result. It shows that the fast residualised scanning approach is algebraically consistent with exact regression for the season-only setup. That is reassuring for the computation.

But this does not validate the underlying modelling assumptions. It confirms that the code is behaving correctly, not that independence, specification, or uncertainty quantification are fully adequate.

```{r}
trait_to_check <- "Lint_percentage"

top20_markers <- scan_season_only_adj %>%
  dplyr::filter(trait == trait_to_check) %>%
  dplyr::arrange(p.value) %>%
  dplyr::slice_head(n = 20) %>%
  dplyr::pull(SNP_ID)
```

```{r}
# ---- Create season-residualised traits ----

make_season_residuals <- function(df, trait_names) {
  df <- as.data.frame(df)
  
  for (tr in trait_names) {
    stopifnot(tr %in% names(df))
    
    tmp <- data.frame(
      .y = df[[tr]],
      .season = df[["Season"]]
    )
    
    keep <- stats::complete.cases(tmp)
    resid_vec <- rep(NA_real_, nrow(df))
    
    fit <- stats::lm(.y ~ .season, data = tmp[keep, , drop = FALSE])
    resid_vec[keep] <- stats::resid(fit)
    
    df[[paste0(tr, "_season_resid")]] <- resid_vec
  }
  
  df
}

analysis_df_resid <- make_season_residuals(analysis_df, trait_names)

# optional quick check
colnames(analysis_df_resid)[grepl("_season_resid$", colnames(analysis_df_resid))]
```

```{r}
stopifnot(exists("analysis_df_resid"))
stopifnot(is.data.frame(analysis_df_resid))
stopifnot(all(paste0(trait_names, "_season_resid") %in% names(analysis_df_resid)))
```

```{r}
exact_resid_lm_results <- run_exact_resid_lm(
  df = analysis_df_resid,
  resid_trait = paste0(trait_to_check, "_season_resid"),
  marker_names = top20_markers
)

knitr::kable(exact_resid_lm_results, digits = 4)
```

```{r}
compare_resid_vs_scan <- scan_season_only_adj %>%
  dplyr::filter(trait == trait_to_check, SNP_ID %in% top20_markers) %>%
  dplyr::select(
    SNP_ID,
    estimate_scan = estimate,
    p_scan = p.value,
    log10p_scan = log10p
  ) %>%
  dplyr::left_join(
    exact_resid_lm_results %>%
      dplyr::select(
        SNP_ID,
        estimate_resid_lm = estimate,
        p_resid_lm = p.value
      ),
    by = "SNP_ID"
  )

knitr::kable(compare_resid_vs_scan, digits = 4)
```

```{r}
run_lasso_trait_checked <- function(df, trait, marker_names,
                                    covariates = c("Season", "PC1", "PC2", "PC3", "PC4", "PC5"),
                                    alpha = 1,
                                    nfolds = 10,
                                    seed = 123) {
  df <- as.data.frame(df)
  
  stopifnot(trait %in% names(df))
  stopifnot(all(covariates %in% names(df)))
  stopifnot(all(marker_names %in% names(df)))
  
  keep <- complete.cases(df[, c(trait, covariates), drop = FALSE])
  df <- df[keep, , drop = FALSE]
  
  y <- df[[trait]]
  X_marker <- prep_marker_matrix(df, marker_names)
  
  X_cov <- model.matrix(
    stats::reformulate(covariates),
    data = df
  )[, -1, drop = FALSE]
  
  x_all <- cbind(X_cov, X_marker)
  cov_names <- colnames(X_cov)
  
  penalty_factor <- c(rep(0, ncol(X_cov)), rep(1, ncol(X_marker)))
  
  stopifnot(length(penalty_factor) == ncol(x_all))
  
  set.seed(seed)
  cvfit <- glmnet::cv.glmnet(
    x = x_all,
    y = y,
    family = "gaussian",
    alpha = alpha,
    nfolds = nfolds,
    standardize = TRUE,
    intercept = TRUE,
    penalty.factor = penalty_factor
  )
  
  coef_min <- coef(cvfit, s = "lambda.min")
  coef_1se <- coef(cvfit, s = "lambda.1se")
  
  nz_min <- rownames(coef_min)[as.vector(coef_min != 0)]
  nz_1se <- rownames(coef_1se)[as.vector(coef_1se != 0)]
  
  nz_min <- setdiff(nz_min, "(Intercept)")
  nz_1se <- setdiff(nz_1se, "(Intercept)")
  
  covariates_nonzero_min <- intersect(nz_min, cov_names)
  covariates_nonzero_1se <- intersect(nz_1se, cov_names)
  
  selected_markers_min <- setdiff(nz_min, cov_names)
  selected_markers_1se <- setdiff(nz_1se, cov_names)
  
  tibble::tibble(
    trait = trait,
    lambda_min = cvfit$lambda.min,
    lambda_1se = cvfit$lambda.1se,
    n_covariates_nonzero_min = length(covariates_nonzero_min),
    n_covariates_nonzero_1se = length(covariates_nonzero_1se),
    n_selected_markers_min = length(selected_markers_min),
    n_selected_markers_1se = length(selected_markers_1se),
    covariates_nonzero_min = list(covariates_nonzero_min),
    covariates_nonzero_1se = list(covariates_nonzero_1se),
    selected_markers_min = list(selected_markers_min),
    selected_markers_1se = list(selected_markers_1se),
    cvfit = list(cvfit)
  )
}
```

```{r}
lasso_results_checked <- lapply(trait_names, function(tr) {
  run_lasso_trait_checked(
    df = analysis_df,
    trait = tr,
    marker_names = marker_names,
    covariates = c("Season", "PC1", "PC2", "PC3", "PC4", "PC5"),
    alpha = 1,
    nfolds = 10,
    seed = 123
  )
}) %>%
  dplyr::bind_rows()

lasso_results_checked %>%
  dplyr::select(
    trait,
    lambda_min, lambda_1se,
    n_covariates_nonzero_min, n_covariates_nonzero_1se,
    n_selected_markers_min, n_selected_markers_1se
  ) %>%
  knitr::kable(digits = 3)
```

```{r}
lasso_cov_diag <- lasso_results_checked %>%
  dplyr::select(trait, covariates_nonzero_min, covariates_nonzero_1se)

knitr::kable(
  lasso_cov_diag %>%
    dplyr::mutate(
      covariates_nonzero_min = purrr::map_chr(covariates_nonzero_min, ~ paste(.x, collapse = ", ")),
      covariates_nonzero_1se = purrr::map_chr(covariates_nonzero_1se, ~ paste(.x, collapse = ", "))
    ),
  digits = 3
)
```

```{r}
lasso_selected_min <- lasso_results_checked %>%
  dplyr::select(trait, selected_markers_min) %>%
  tidyr::unnest_longer(selected_markers_min, values_to = "SNP_ID") %>%
  dplyr::filter(!is.na(SNP_ID))

lasso_selected_1se <- lasso_results_checked %>%
  dplyr::select(trait, selected_markers_1se) %>%
  tidyr::unnest_longer(selected_markers_1se, values_to = "SNP_ID") %>%
  dplyr::filter(!is.na(SNP_ID))
```

```{r}
lasso_results_final <- lapply(trait_names, function(tr) {
  run_lasso_trait_checked(
    df = analysis_df,
    trait = tr,
    marker_names = marker_names,
    covariates = c("Season"),
    alpha = 1,
    nfolds = 10,
    seed = 123
  )
}) %>%
  dplyr::bind_rows()

lasso_results_final %>%
  dplyr::select(
    trait,
    lambda_min, lambda_1se,
    n_covariates_nonzero_min, n_covariates_nonzero_1se,
    n_selected_markers_min, n_selected_markers_1se
  ) %>%
  knitr::kable(digits = 3)
```

```{r}
lasso_final_selected_min <- lasso_results_final %>%
  dplyr::select(trait, selected_markers_min) %>%
  tidyr::unnest_longer(selected_markers_min, values_to = "SNP_ID") %>%
  dplyr::filter(!is.na(SNP_ID))

lasso_final_selected_1se <- lasso_results_final %>%
  dplyr::select(trait, selected_markers_1se) %>%
  tidyr::unnest_longer(selected_markers_1se, values_to = "SNP_ID") %>%
  dplyr::filter(!is.na(SNP_ID))
```

The report uses cross-validated `glmnet` with `alpha = 1`, leaving season and PC covariates unpenalised. Under the **Season + PC1–5** specification, `lambda.min` selects roughly **212–240 markers** and `lambda.1se` roughly **140–190 markers**, while all seven covariates remain non-zero. Under the **Season only** specification, `lambda.min` still selects roughly **220–247 markers**, and `lambda.1se` still keeps around **161–181 markers**.

This is highly informative: **even after shrinkage, the selected marker set remains large**. That implies the signal is not extremely sparse. Instead, the architecture looks more like a combination of correlated marker sets, block-level signal, and/or moderately diffuse genetic effects.

There are important interpretive limits, though. Under strong predictor correlation, LASSO often chooses representatives from correlated groups rather than identifying true causal variables. So the selected set is better understood as a **predictive subset** than a causal one. Also, the report does not place LASSO alongside ridge or elastic net, so the multi-marker shrinkage comparison is still incomplete.

```{r}
trait_to_plot <- "Lint_percentage"

top_snp <- scan_season_pc_adj %>%
  dplyr::filter(trait == trait_to_plot) %>%
  dplyr::arrange(p.value) %>%
  dplyr::slice(1) %>%
  dplyr::pull(SNP_ID)

top_snp

analysis_df %>%
  dplyr::mutate(
    genotype_class = factor(.data[[top_snp]])
  ) %>%
  ggplot(aes(x = genotype_class, y = .data[[trait_to_plot]], fill = genotype_class)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.15, alpha = 0.35, size = 1) +
  facet_wrap(~ Season) +
  labs(
    title = paste("Phenotype by genotype class:", top_snp),
    x = paste("Genotype at", top_snp),
    y = trait_to_plot
  ) +
  theme_minimal(base_size = 12)
```

`Marker-4017` emerges as the top SNP for `Lint_percentage` under the season+PC model. The boxplots by genotype class show clear trait separation across genotype classes within seasons. After residualising `Lint_percentage` on `Season + PC1–PC5`, the genotype-class differences remain visible in the residual plot.

This is a strong robustness check at the visual level. It indicates that the `Marker-4017` signal is not merely an artefact of seasonal shift or broad genotype structure. Even after removing those broad covariate effects, the genotype classes retain clear phenotypic separation. That said, the proper conclusion is still that `Marker-4017` is a strong **candidate signal region**, not necessarily a causal variant by itself.

### 

```{r}
trait_to_plot <- "Lint_percentage"

top_snp <- scan_season_pc_adj %>%
  dplyr::filter(trait == trait_to_plot) %>%
  dplyr::arrange(p.value) %>%
  dplyr::slice(1) %>%
  dplyr::pull(SNP_ID)

plot_df <- analysis_df %>%
  dplyr::filter(
    !is.na(.data[[trait_to_plot]]),
    !is.na(.data[[top_snp]])
  ) %>%
  dplyr::mutate(
    trait_resid = resid(
      lm(
        as.formula(
          paste(trait_to_plot, "~ Season + PC1 + PC2 + PC3 + PC4 + PC5")
        ),
        data = .
      )
    ),
    genotype_class = factor(.data[[top_snp]])
  )

ggplot(plot_df, aes(x = genotype_class, y = trait_resid, fill = genotype_class)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.15, alpha = 0.35, size = 1) +
  labs(
    title = paste("Season/PC-adjusted", trait_to_plot, "by", top_snp),
    x = paste("Genotype at", top_snp),
    y = paste(trait_to_plot, "(residualised)")
  ) +
  theme_minimal(base_size = 12)
```

```{r}
library(tidyverse)
library(patchwork)

scan_plot_df <- scan_season_pc_adj
# 如果想画 season-only 版本，改成：
# scan_plot_df <- scan_season_only_adj

traits_to_plot <- c(
  "Lint_percentage",
  "Seed_index",
  "Lint_index",
  "Seed_oil_content"
)

chr_levels <- unique(map_df$Chromosome)

genome_map <- map_df %>%
  dplyr::filter(!is.na(Chromosome), !is.na(Position_bp)) %>%
  dplyr::mutate(
    Chromosome = factor(Chromosome, levels = chr_levels)
  ) %>%
  dplyr::arrange(Chromosome, Position_bp)

chr_df <- genome_map %>%
  dplyr::group_by(Chromosome) %>%
  dplyr::summarise(chr_len = max(Position_bp), .groups = "drop") %>%
  dplyr::mutate(offset = dplyr::lag(cumsum(chr_len), default = 0))

genome_map2 <- genome_map %>%
  dplyr::left_join(chr_df, by = "Chromosome") %>%
  dplyr::mutate(pos_cum = Position_bp + offset)

axis_df <- genome_map2 %>%
  dplyr::group_by(Chromosome) %>%
  dplyr::summarise(center = mean(range(pos_cum)), .groups = "drop")

plot_manhattan_trait <- function(trait_name, scan_df, genome_map2, axis_df) {

  df_plot <- scan_df %>%
    dplyr::filter(trait == trait_name) %>%
    dplyr::select(SNP_ID, p.value, log10p, Chromosome, Position_bp) %>%
    dplyr::filter(
      !is.na(p.value),
      !is.na(Chromosome),
      !is.na(Position_bp)
    ) %>%
    dplyr::mutate(
      Chromosome = factor(Chromosome, levels = levels(genome_map2$Chromosome))
    ) %>%
    dplyr::left_join(
      genome_map2 %>% dplyr::select(SNP_ID, pos_cum),
      by = "SNP_ID"
    ) %>%
    dplyr::arrange(Chromosome, Position_bp)

  # Bonferroni line
  bonf_line <- -log10(0.05 / nrow(df_plot))

  ggplot(df_plot, aes(x = pos_cum, y = log10p, colour = Chromosome)) +
    geom_point(alpha = 0.75, size = 0.75) +
    geom_hline(
      yintercept = bonf_line,
      linetype = "dashed"
    ) +
    scale_x_continuous(
      breaks = axis_df$center,
      labels = axis_df$Chromosome
    ) +
    labs(
      title = paste("Manhattan plot:", trait_name),
      x = "Chromosome",
      y = expression(-log[10](p))
    ) +
    theme_minimal(base_size = 12) +
    theme(
      legend.position = "none",
      panel.grid.minor.x = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
}

p1 <- plot_manhattan_trait("Lint_percentage",   scan_plot_df, genome_map2, axis_df)
p2 <- plot_manhattan_trait("Seed_index",        scan_plot_df, genome_map2, axis_df)
p3 <- plot_manhattan_trait("Lint_index",        scan_plot_df, genome_map2, axis_df)
p4 <- plot_manhattan_trait("Seed_oil_content",  scan_plot_df, genome_map2, axis_df)

p1
p2
p3
p4

```

The Manhattan plots display genome-wide significance profiles for `Lint_percentage`, `Seed_index`, `Lint_index`, and `Seed_oil_content`, with Bonferroni thresholds drawn as dashed lines. Across traits, the plots show **multiple elevated regions rather than isolated single spikes**. The peaks for `Lint_percentage` and `Lint_index` appear especially prominent, while `Seed_index` and `Seed_oil_content` also show multiple threshold-exceeding regions.

These plots are fully consistent with the earlier count summaries: the signal landscape is **broad and clustered**, not sparse and isolated. That pattern strongly suggests LD-driven blocks and/or polygenic structure rather than a tiny number of independent loci.

This also explains why LASSO continues to retain well over a hundred markers: the Manhattan plots themselves show that the genome-wide signal pattern is regionally distributed rather than point-like.

### 
